---
title: "Predicting Success for Olympic Track Athletes with a Multi-Level Model"
author: "Nick Browen, Eric Ortiz"
output: html_notebook
indent: true
---

```{r echo=F}
pacman::p_load(tidyverse, nlme, lme4)
track <- read.csv("trackFinal.csv")
```

```{r}
track
```

=============================================================
  
  Project Description:
  
  http://statweb.calpoly.edu/bchance/stat414/Stat414Project.pdf

=============================================================

### Introduction 

```
Still to do: 
  1. Incorporate supporting references into text
```
  Every four years, the Olympics Events garner worldwide attention much attention. We wanted to know what athlete characteristics contribute to their success and how the country they are from affects this. There is so much data readily available about the Olympics, even going back to 1896. We also were excited to be able to combine datasets and information from multiple resources from Olympic data to population and GDP of the countries of the world over time.  
  
  We chose to include athlete-level characteristics because the ideal physique for a distance runner is much different than a 100m sprinter and we think this will be useful in explaining differences in finishing time and are curious what other nuances will be revealed. At the country-level, we chose to include explanatory variables because athletes are sent as a team by their country and so the athlete pool, training resources and even quality of life might reveal some trends in predicting finishing times.  
  
  When looking through some literature on the subject, we came to realize that more is involved with success in the Olympics than just the variables at the athlete level such as height, weight, gender, and age. We were struck by Xun Bian's paper titled "Predicting Olympic Medal Counts: the Effects of Economic Development on Olympic Performance" in which Olympic medal counts for a country were predicted from variables at the country level such as population, GDP, who the hosting country is, and whether the country is Socialist or not. Further, when we looked at the paper by Filippo Radicchi titled "Universality, Limits and Predictability of Gold-Medal Performances at the Olympic Games", we could see that there is variablility at the athlete level since many athletes compete in more than one Olympics in their lifetime.
  
  We chose to consider factors at both the athlete level and country level in a Hierarchical Model to predict the finishing time of track athletes.
  
  
##### Research Question
How can we best predict an Olympic track athlete’s finishing time? What is the relationship between an athlete’s finishing time and factors at the athlete level such as sex, age, weight, and height as well as factors at the country level such as the athlete’s nationality, their country's GDP, and population?
  
  
  **Chance's Notes**: 
  *I really like how you have athlete level variables and country level variables.  But I do wonder if you need to “standardize” the response variable a bit?  Nesting within event is good but still difficult to compare the times to each other?  Would it be better to pick one event and look at multiple athletes from the same country over time?*



### Materials and Methods 
```
Three to five paragraphs (or fewer) that…
  1. Briefly describe your data, where it came from (source), definitions of important variables, 
     and how it was collected
  2. Indicate any modifications made to the data, recoding, or decisions about missing data
  3. Briefly but thoroughly describe the statistical inference methods used to quantify the 
     association between your outcome and predictor variables in a multilevel analysis. 
     What summary statistics were calculated? 
     What statistical tests were performed?
  4. Specify strategies employed when building your models
  5. Do not report results in the Materials and Methods section! 
```
Our data consist of merged datasets from Kaggle.com. First we found a comprehensive dataset of all Olympic medalists, but we subsetted this to include just track athletes (specifically the 10k race and all events shorter than 10k). This file contains info on the athlete level such as height, weight, and age, but it did not include the finishing times. So we then merged this with another dataset from Kaggle that included finishing time. Then, from Gapminder.com we obtained country level information such as GDP and population and merged this into our dataset as well.


### Results
```
The meat of your report, which should include…
  1. A general description of your data (completed via your exploratory data analysis)
  2. A description of the results from your analyses, including interpretations of parameter estimates, 
     tests, and confidence intervals in context.
  3. Tables that summarize results and figures that illustrate results. These tables and figures should 
     be well-labeled, numbered (e.g., Figure 1), and have a good, descriptive caption. Each report 
     should have a minimum of two plots; rarely are residual plots part of the main body of the report 
     unless they are an integral part of the story.
  4. You should interpret tests, confidence intervals, and coefficients in this section, but you should 
     not editorialize here! Save that for the Discussion. 
```

```{r}
mod13 <- lme(data = track, 
            fixed = timeSecs ~ dist100 + c_BMI + year1896 + sex + sex*dist100 + gdpPerCap_ + gdpPerCap_*dist100,
            random = ~ dist100|country2)
summary(mod13)
```



### Discussion
```
A few paragraphs that:
  1. Begin with an accurate summary statement; describe how the results help answer your research questions 
     and what was most interesting from your analysis. In fact, the first paragraph of the Discussion is 
     very important – in professional journals, it is often the first and sometimes the only paragraph that 
     is read in a paper. After the first sentence highlights primary results, the remainder of the first 
     paragraph might compare your results to others in the literature or include interesting secondary 
     results.
  2. Discuss possible implications of the results in the context of the research question.
  3. Make a statement regarding potential confounding variables in your study.
  4. Make a statement about the generalizability of your results. Don’t give generic statements of possible 
     causation and generalizability, but thoughtfully discuss relevant issues – confounding variables, 
     representativeness of the sample, etc.
  5. Identify any limitations of your study. Discuss the potential impact of such limitations on the 
     conclusions.
  6. Identify strengths and weaknesses of your analysis.
  7. Make suggestions for future research. Identify important next steps that a researcher could take to 
     build on your work.
  8. Do not include test statistics or p-values in this section. 
```


### Annotated Appendix
```
  1. Tables and figures that are informative but were not referenced specifically in the main report. 
     Include a short annotation – one or two sentences on what they show.
  2. Annotated output so that I can trace how you constructed your final data set, what models you ran to 
     produce the results quoted in your report, and what intermediate models you also considered.
  3. Description of statistical modeling steps that were not included in the main body of your report. 
     Possible entries here include:
      - How you handled missing data
      - Evaluation of assumptions.
      - Outlier analysis and how you decided to deal with any outliers along with rationale for your decision.
      - Describe hypotheses testing you performed during model building and how you decided on the explanatory 
        variables you ultimately included in your final model. 
      - Assessment of the final model.
  4. How you went from the model output in JMP/R to interpretations in your report 
     (e.g. exponentiate coefficients, and then take inverse)
  5. Anticipate questions someone might have after reading your report, and make sure those questions can be 
     answered with information in the appendix.
  6. A citation for each reference article you included in your proposal. Also include a link, if appropriate. 
     Remember that you must have the entire paper and not just an abstract, and at least two must be from peer-
     reviewed journals. 
```

```{r echo=FALSE}
mod13_diagnosics <- data.frame(fittedValues = fitted.values(mod13), residuals = residuals(mod13))
ggplot(mod13_diagnosics, aes(x=fittedValues, y=residuals)) + 
  geom_point(size = 0.5) +
  geom_hline(yintercept = 0, color = "red") + 
  labs(title = "Residuals vs Fitted",
       subtitle = "Model 13")
```


```{r echo=FALSE}
ggplot(mod13_diagnosics) + 
  geom_qq_line(aes(sample=residuals), color = "red") +
  geom_qq(aes(sample=residuals), size = 0.5) +
  labs(title = "QQ-Plot",
       subtitle = "Model 13",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles")
```



```{r echo=FALSE}
ranef_mod13 <- ranef(mod13) %>% as.data.frame()
colnames(ranef_mod13) <- c("intercept", "dist100")

ggplot(data=ranef_mod13, aes(intercept)) + 
  geom_histogram(bins = 20) +
  labs(title = "Histogram of Random Intercepts",
       subtitle = "Model 13",
       y = "Count",
       x = "Intercepts")
```

```{r echo=FALSE}
ggplot(data=ranef_mod13, aes(dist100)) + 
  geom_histogram(bins = 15) + 
  ggtitle("Random Slopes for mod13") +
  labs(title = "Histogram of Random Slopes",
       subtitle = "Model 13",
       x = "Slopes",
       y = "Count")
```

##### Citations
1. Radicchi F (2012) Universality, Limits and Predictability of Gold-Medal Performances at the Olympic Games.
  https://doi.org/10.1371/journal.pone.0040335 
   
2. Bian, X. 2005. Predicting Olympic Medal Counts: The Effects of Economic Development on Olympic Performance.
  https://pdfs.semanticscholar.org/7293/1ab692bcab9e724b0e5ed4adb53b7ff8097f.pdf
  
3. Country Level Data such as GDP and Population.
  https://www.gapminder.org/data/
  
4. Athlete Level Data for Finishing Time.
  https://www.kaggle.com/jayrav13/olympic-track-field-results
  
5. Athlete Level Data for Height, Age, Sex, and Country.
  https://www.kaggle.com/heesoo37/120-years-of-olympic-history-athletes-and-results