---
title: "EDA Output"
output: html_notebook
---

Data
```{r}
track0 <- read.csv("C:/Users/nickbrowen/Documents/GitHub/STAT-414-Project/data files/final_data_noNA3.csv")
track0 <- track0[,-1]
track0$medal <- ordered(track0$medal, levels = c("B", "S", "G"))
```

Level 1 = Athlete (finishing time)
Level 2 = Country
Level 3 = Event


```{r}
track <- track0 %>% filter(event != "110M Hurdles Men"& event != "3000M Steeplechase Men" & event != "400M Hurdles Men" & event != "100M Hurdles Women" & event != "3000M Steeplechase Women" & event !="400M Hurdles Women")
track$dist <- as.numeric(gsub("[^0-9]", "",track$event))

write.csv(track, file = "track_dist.csv")
```

```{r}
track$country <- factor(track$country)
track$country2 <- factor(track$country2)
```


Level 1 = Athlete (finishing time)
  vars: name, event, timeSecs, timeMins, year (?), medal, sex, height, weight, age, dist 
Level 2 = Country
  vars: country, country2, gdp, pop
  
(?) - should year be its own level? i.e. athlete in year in country
better to be athlete in country but I'm not sure if what to do with year

#Response
```{r}
track %>% select(timeSecs, timeMins) %>% summary
```



##Level 1 (Athlete)

Numerical Summary for Athlete Level (L1) Quantitative Variables
```{r}
L1 <- c("name", "sex", "age", "height", "weight", "medal")
L1_qvar <- c("age", "height", "weight")
track %>% select(L1_qvar) %>% summary
```

```{r}
tapply(track$timeSecs, track$dist, mean)
```



```{r}
ggplot(data = track, aes(age)) + geom_histogram(bins = 25)
```
```{r}
ggplot(data = track, aes(height)) + geom_histogram(bins = 25)
```
```{r}
ggplot(data = track, aes(weight)) + geom_histogram(bins = 30)
```
```{r}
library(corrplot)
L1corr_matrix <- track %>% select(L1_qvar, timeSecs, year, dist) %>% cor()
corrplot(L1corr_matrix, type = "upper")
```
```{r}
L1corr_matrix
```



```{r}
ggplot(data = track, aes(x=height, y = timeSecs)) + geom_point() + stat_smooth(method = "lm") + scale_y_continuous(limits = c(0, 2000))
```


```{r}
ggplot(data = track, aes(x=weight, y = timeSecs)) + geom_point() + stat_smooth(method = "lm") + scale_y_continuous(limits = c(0, 2000))
```
```{r}
ggplot(data = track, aes(x=height, y = weight)) + geom_point()
```
```{r}
ggplot(data = track, aes(x=year, y=age)) + geom_point() + stat_smooth(method="lm")
```


```{r}
ggplot(data = track, aes(x=dist, y=timeMins)) + geom_point() + stat_smooth(method="lm")
```
```{r}
ggplot(data=track %>% filter(timeMins < .5)) + geom_point(aes(x = year, y=timeMins, color = event)) +   geom_line(aes(x=year, y=timeMins, color=event))
```

```{r}
ggplot(data=track %>% filter(timeMins > .5 & timeMins<10)) + geom_point(aes(x = year, y=timeMins, color = event)) +   geom_line(aes(x=year, y=timeMins, color=event))
```


```{r}
ggplot(data = track, aes(as.factor(dist), fill =sex)) + geom_bar() + xlab("Distance") 
```

```{r}
ggplot(data = track, aes(group=as.factor(sex), y=timeMins)) + geom_boxplot() + facet_wrap(aes(dist), scales = "free") + xlab("Sex")
```


```{r}
ggplot(data = track, aes(sex, fill = sex)) + geom_bar(width = 0.5)
```

Possible Level 1 Interactions

```{r}
ggplot(data = track, aes(x=height, y=timeSecs)) + geom_point() + facet_wrap(aes(dist), scales = "free") + stat_smooth(method ="lm")
```
slopes all seem similiar, no interaction with distance and height


```{r}
ggplot(data = track, aes(x=weight, y=timeSecs)) + geom_point() + facet_wrap(aes(dist), scales = "free") + stat_smooth(method ="lm")
```
also no visual evidence of interaction with weight and distance

```{r}
ggplot(data = track, aes(x=age, y=timeSecs)) + geom_point() + facet_wrap(aes(dist), scales = "free") + stat_smooth(method ="lm")
```
differences in directions of slope, visual evidence for interaction with distanc and age

```{r}
ggplot(data = track, aes(x=year, y=timeSecs)) + geom_point() + facet_wrap(aes(dist), scales = "free") + stat_smooth(method ="lm")
summary(lm(track$timeSecs ~ track$year + track$dist))
```



```{r}
scale_dist <- tapply(track$timeMins, track$dist, scale)
scale_country <- tapply(track$timeMins, track$country2, scale)

boxplot(scale_dist, main = "Comparing Spread of Time (Group Scaled) by Distance")
```

```{r}
tapply(track$timeMins, track$dist, sd)
```
```{r}
ggplot(data = track, aes(y=timeSecs)) + geom_boxplot() + facet_wrap(aes(dist), scales="free") 
```





##Level 2 (Country)

```{r}
sort(table(track$country), desc = T)
```



```{r}
ggplot(data = track, aes(country, fill = medal)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```



Numerical Summary for Country Level (L2) Quantitative Variables
```{r}
L2 <- c("country2", "gdp", "pop")
L2_qvar <- c("gdp", "pop")
track %>% select(L2_qvar) %>% summary
```
```{r}
ggplot(data = track, aes(gdp)) + geom_histogram(bins = 30)
```
```{r}
ggplot(data = track, aes(pop)) + geom_histogram(bins = 30)
```
```{r}
track %>% filter(pop > 500000000) 
```
Will have to keep in mind that China is an extreme population outlier and they have 4 observations in the dataset.

```{r}
L2corr_matrix <- track %>% select(L2_qvar, timeSecs) %>% cor()
corrplot(L2corr_matrix, type = "upper")
```
```{r}
L2corr_matrix
```


```{r}
ggplot(data = track %>% subset(pop<1000000000), aes(x=pop, y=timeMins)) + geom_point() + stat_smooth(method = "lm") + facet_wrap(aes(dist), scales = "free") + ggtitle("Time vs Pop by Dist excluding China")
```

```{r}
track$logpop <- log(track$pop)
ggplot(data = track, aes(x=logpop, y=timeMins)) + geom_point() + stat_smooth(method = "lm") + facet_wrap(aes(dist), scales = "free") + ggtitle("Time vs Log(Pop) by Dist including China")
```



```{r}
ggplot(data = track, aes(x=gdp, y=timeMins)) + geom_point() + stat_smooth(method = "lm") + facet_wrap(aes(dist), scales = "free")
```

```{r}
ggplot(data = track, aes(x=logpop, y=gdp)) + geom_point()+ stat_smooth(method = "lm")
```



Distribution of Medals by Country
```{r}
t <- track %>% select(c("country2", "medal")) %>% table() 
addmargins(t,margin = seq_along(dim(t)), FUN = sum)

```




```{r}
countrysd <- tapply(track$timeMins, track$country, var, simplify = T)
countrysd_noNA <- countrysd[!is.na(countrysd) & !is.null(countrysd)]

```

```{r}
rownames(countrysd_noNA) <- NULL
barplot(countrysd_noNA, xlab = "Countrys", ylab = "Standard Deviation (TimeMins)")
```



```{r}
track %>% filter(pop < 500000000) %>% ggplot(aes(y=timeSecs, x=pop, color = event)) + geom_point() +   geom_smooth(method = "lm", se = F)
```


```{r}
track %>% filter(pop < 500000000) %>% ggplot(aes(y=gdp, x=pop, color = medal)) + geom_point() + geom_smooth(method = "lm", se = F)
```


```{r}
tapply(track$gdp, track$medal, mean)
tapply(track$pop, track$medal, mean)       
```


```{r}
library(corrplot)
corr_matrix <- track %>% select(contin_var, dist, logpop) %>% cor()
corrplot(corr_matrix, type = "upper")
```
```{r}
corr_matrix
```




