---
title: "Merge"
output: html_notebook
---

```{r}
library(tidyverse)
library(reshape2)
library(lubridate)
```


```{r}
olympics <- read.csv("Track-Results.csv")
pop <- read.csv("population.csv")
gdp <- read.csv("gdp.csv")
codes <- read.csv("country-codes.csv")
```

```{r}
head(olympics)
head(pop)
head(gdp)
head(codes)
```

```{r}
codes
```





goal: long data format
Geo   Year    GDP
---------------------
Afg   1800    3280000
Afg   1801    3280000
...

```{r}
# Convert gdp and pop data into long format
pop_long <- pop %>% 
  melt() %>% 
  mutate(year = variable %>% substr(2,5) %>% as.numeric()) %>%
  select(geo, year, value)

colnames(pop_long) <- c("country", "year", "population")

gdp_long <- gdp %>% 
  melt() %>%
  mutate(year = variable %>% substr(2,5) %>% as.numeric()) %>%
  select(geo, value, year)

colnames(gdp_long) <- c("country", "gdp", "year")

# Join the two data sets
country_data <- full_join(gdp_long, pop_long, by=c("country", "year")) %>%
  select(year, country, gdp, population)
```


```{r}
colnames(codes) <- c("x", "country", "code")
codes <- codes %>% select(code, country)
codes$code <- codes$code %>% as.character()
codes$country <- codes$country %>% as.character()
```


USA
GBR
```{r}
codes
```


do a join on x=country data, y=codes to get the country codes and keep all observations in x

```{r}
country_codes <- full_join(country_data, codes, by="country")

country_codes %>% filter(year == 1801) %>% filter(is.na(code)) %>% select(year, country)

codes %>% filter(country == "Bolivia")
```








Now merge country_codes with olympics.

goal: keep each row in olympics, include that athlete's country, gdp, population

gender  event location  year  medal name  nationality result  gdp, population

```{r}
olympics
colnames(olympics) <- c("gender", "event", "hostCity", "year", 
                        "medal", "name", "code", "result", "X")
olympics <- olympics %>% select(name, code, year, event, medal, result, hostCity, gender)
```

```{r}
data <- full_join(olympics, country_codes, by=c("year", "code"))
```
Events to filter out:

Heptathlon Women
Decathlon Men
Discus Throw Men
Hammer Throw Men
High Jump Men
Javelin Throw Men
Long Jump Men
Pole Vault Men
Shot Put Men
Triple Jump Men
Discus Throw Women
Hammer Throw Women
High Jump Women
Javelin Throw Women
Long Jump Women
 Pole Vault Women
 Shot Put Women
 Triple Jump Women
```{r}
data
```


```{r}
drop_events <- c("Heptathlon Women", "Decathlon Men", "Discus Throw Men", "Hammer Throw Men", "High Jump Men", "Javelin Throw Men", 
                 "Long Jump Men", "Pole Vault Men", "Shot Put Men", "Triple Jump Men", "Discus Throw Women", "Hammer Throw Women", 
                 "High Jump Women", "Javelin Throw Women", "Long Jump Women", "Pole Vault Women", "Shot Put Women", "Triple Jump Women",
                 "Decathlon Men", "Discus Throw Men", "Discus Throw Women", "Hammer Throw Men", "Hammer Throw Women", "Heptathlon Women", 
                 "High Jump Men", "High Jump Women", "Javelin Throw Men", "Javelin Throw Women", "Long Jump Men", "Long Jump Women",
                 "Pole Vault Men", "Pole Vault Women", "Shot Put Men", "Shot Put Women", "Triple Jump Men", "Triple Jump Women",
                 "20Km Race Walk Men", "50Km Race Walk Men", "Marathon Men", "20Km Race Walk Women", "Marathon Women")

'%!in%' <- function(x,y)!('%in%'(x,y))

track_data <- data %>% filter(event %!in% drop_events) %>%
  filter(!is.na(event))

```

```{r}
track_data$event %>% unique()
```


```{r}
track_data$name <- track_data$name %>% tolower()
track_data$event %>% unique()
```


```{r}
medalists <- read.csv("track_medalists.csv")
medalists$Name <- medalists$Name %>% tolower()
```

```{r}
medalists <- medalists %>% mutate(
  Medal = case_when(Medal == "Gold" ~ "G",
            Medal == "Silver" ~ "S",
            Medal == "Bronze" ~ "B")) %>%
  select(-one_of("X"))
```


```{r}
#Pare this down to the 29 events in track data
keep.events <- c("Athletics Men's 10,000 metres", "Athletics Men's 100 metres", 
                 "Athletics Men's 110 metres Hurdles", "Athletics Men's 1,500 metres", 
                 "Athletics Men's 200 metres", "Athletics Men's 3,000 metres Steeplechase", 
                 "Athletics Men's 400 metres Hurdles", "Athletics Men's 400 metres", 
                 "Athletics Men's 4 x 100 metres Relay", "Athletics Men's 4 x 400 metres Relay", 
                 "Athletics Men's 5,000 metres", "Athletics Men's 800 metres",
                 "Athletics Women's 10,000 metres", "Athletics Women's 100 metres Hurdles", 
                 "Athletics Women's 100 metres", "Athletics Women's 10,000 metres", 
                 "Athletics Women's 200 metres", "Athletics Women's 3,000 metres Steeplechase", 
                 "Athletics Women's 400 metres Hurdles", "Athletics Women's 400 metres",
                 "Athletics Women's 4 x 100 metres Relay", "Athletics Women's 4 x 400 metres Relay", 
                 "Athletics Women's 5,000 metres", "Athletics Women's 800 metres", "Athletics Women's 1,500 metres" )

medalists <- medalists %>% filter(Event %in% keep.events)
```

```{r}
colnames(medalists) <- c("name", "year", "sex", "age", "height", "weight", "country", "code",
                         "event", "medal")
medalists <- medalists %>% select(name, event, year, medal, code, country, sex, age, height, weight, country)
```

```{r}
medalists <- medalists %>% mutate(
  event = case_when(event == "Athletics Men's 10,000 metres" ~ "10000M Men",
                    event == "Athletics Men's 100 metres" ~ "100M Men",
                    event == "Athletics Men's 110 metres Hurdles" ~ "110M Hurdles Men",
                    event == "Athletics Men's 1,500 metres" ~ "1500M Men",
                    event == "Athletics Men's 200 metres" ~ "200M Men",                 
                    event == "Athletics Men's 3,000 metres Steeplechase" ~ "3000M Steeplechase Men",  
                    event == "Athletics Men's 400 metres Hurdles" ~ "400M Hurdles Men",       
                    event == "Athletics Men's 400 metres" ~ "400M Men",                 
                    event == "Athletics Men's 4 x 100 metres Relay" ~ "4X100M Relay Men",     
                    event == "Athletics Men's 4 x 400 metres Relay" ~ "4X400M Relay Men",       
                    event == "Athletics Men's 5,000 metres" ~ "5000M Men",               
                    event == "Athletics Men's 800 metres" ~ "800M Men",                
                    event == "Athletics Women's 10,000 metres" ~ "10000M Women",
                    event == "Athletics Women's 100 metres Hurdles" ~ "100M Hurdles Women",
                    event == "Athletics Women's 100 metres" ~ "100M Women", 
                    event == "Athletics Women's 10,000 metres" ~ "10000M Women",
                    event == "Athletics Women's 200 metres" ~ "200M Women",
                    event == "Athletics Women's 3,000 metres Steeplechase" ~ "3000M Steeplechase Women",
                    event == "Athletics Women's 400 metres Hurdles" ~ "400M Hurdles Women",   
                    event == "Athletics Women's 400 metres" ~ "400M Women",
                    event == "Athletics Women's 4 x 100 metres Relay" ~ "4X100M Relay Women", 
                    event == "Athletics Women's 4 x 400 metres Relay" ~ "4X400M Relay Women",  
                    event == "Athletics Women's 5,000 metres" ~ "5000M Women",     
                    event == "Athletics Women's 800 metres" ~ "800M Women",
                    event == "Athletics Women's 1,500 metres" ~ "1500M Women"))

medalists$medal <- medalists$medal %>% as.factor()
medalists$event <- medalists$event %>% as.factor()
track_data$code <- track_data$code %>% as.factor()
```


year, event, medal, code
```{r}
track_data$event %>% unique()
medalists$event %>% unique()
```

```{r}
joined_data <- full_join(track_data, medalists, by = c("code", "event", "medal", "year")) %>% 
  select(name.x, event, result, code, country.x, year, medal, hostCity, gender, 
        gdp, population, age, height, weight) %>% 
  filter(!is.na(gender))

colnames(joined_data) <- c("name", "event", "time", "country", "country2", "year", "medal", "city", "sex", "gdp",
                           "pop", "age", "height", "weight")
```





```{r}
timeToDecimal <- function(t){
  # t is Time in seconds
  t <- gsub("[^\\d]+", ".", t, perl=TRUE)
    #as.duration()
  lent <- nchar(t)

  if(lent == 1){
    t = NA
  }else if(lent == 2){
    # put in else at end
    t = t
  }else if(lent == 3){
    # else at end
    t = t
  }else if(lent == 4){
    # else at end
    t = t
  }else if(lent == 5){
    t = t
  } else if(lent == 7){
    min <- substr(t, 1, 2) %>% as.numeric() %>% prod(60)
    sec <- substr(t, 4, 7) %>% as.numeric() %>% sum(min)
    t = sec %>% as.character()
  }

  return(t)
}

joined_data$timeSecs <- joined_data$time %>% sapply(timeToDecimal) %>% as.numeric()
joined_data$timeMins <- joined_data$timeSecs / 60

joined_data <- joined_data %>% filter(!is.na(timeMins))

joined_data <- joined_data %>% select(name, event, timeSecs, timeMins, country, country2, year, medal, 
                                      city, sex, gdp, pop, age, height, weight) %>% 
  filter(event %!in% c("4X400M Relay Men", "4X100M Relay Men", "4X100M Relay Women", "4X400M Relay Women"))

na_removed_data <- joined_data %>% filter(!is.na(gdp) & !is.na(age) & !is.na(height) & !is.na(weight))
```


```{r}
joined_data %>% write.csv("final_data2.csv")
na_removed_data %>% write.csv("final_data_noNA2.csv")
```



# Final Datasets
### 1. Joined Data
is the full data set, including missing values

```{r}
joined_data
```

### 2. na_removed_data
Removed the na values for EV's from `joined_data`
```{r}
na_removed_data
```






